---
title: "ENVS-193DS_homework-03"
author: "Aiden Soller"
format: html
editor: visual
---

[GitHub ](https://github.com/aidensoller-glitch/ENVS-193DS_homework-03.git)[repository](https://github.com/aidensoller-glitch/ENVS-193DS_homework-03.git)

```{r}
library(tidyverse)

library(janitor)

library(gt)

library(car)

library(rstatix)

library(dplyr)

library(here)

salinity<- read_csv("../data/salinity-pickleweed.csv") #read in soil salinity data

sleep_calendar <- read_csv("../data/sleep_schedule.csv") # read in project data as an object called sleep_calendar
```

# Problem 1: Slough Soil Salinity

# a. An appropriate test

Two appropriate tests to determine the strength of the relationship are Pearson's product movement correlation and linear regression. Pearson correlation measures both the strength and direction of the linear relationship between soil salinity and pickleweed biomass. Comparatively, linear regression also tests the relationship between salinity and biomass, except it models biomass as the response variable, estimating how much biomass changes with soil salinity.

# b. Create a visualization

```{r}
ggplot(data = salinity, 
       aes(x = salinity_mS_cm, y = pickleweed))+
  geom_point(color = "darkgreen", size = 3) +
  labs(
    x = "Soil Salinity (mS/cm)",
    y = "California Pickleweed Biomass (g)",
    title = "Relationship Between Soil Salinity and Pickleweed Biomass"
  ) +
  theme_minimal()
```

# c. Check your assumptions and run your test

## Checking Assumptions
1. Linear Relationship: Yes, a linear relationship is expected between soil salinity and pickleweed biomass. Factors such as salinity influence plant growth, so a change in soil salinity would correspond to an increase or decrease in biomass.

2. Independent errors: Each pickleweed plant was sampled individually from the slough, so if there is error for one plant it shouldn't influence the error for another.
3. Homoscedasticity: Checking if error is the same throughout range of predictor values
```{r}
model <- lm(pickleweed ~ salinity_mS_cm, data = salinity) # Fit the linear model
plot(model, which = 1) # Residuals vs fitted plot
```
4. Normally Distributed errors
```{r}
plot(model, which = 2) # Q-Q plot of residuals
```

Assumption Description:
I checked the assumptions of linearity, independence of errors, homoscedasticity, and normally distributed errors. Linearity was checked by assessing the potential of a linear relationship between variables, independent errors was evaluated based on sampling design, homoscedasticity was checked using a residuals vs. fitted plot, and normality using a Q-Q plot of the residuals. The scatterplot showed a approximately linear trend between salinity and pickleweed biomass, the residuals vs fitted plot did not show a strong pattern, and the Q-Q plot showed residuals that generally followed the reference line, suggesting that the assumptions of linear models is satisfied.

##Running Test
```{r}
salinity_model <- lm(pickleweed ~ salinity_mS_cm, data = salinity) # Command to run linear regression

summary(salinity_model) # Summary and model displayed
```

# d. Results communication

To evaluate the strength of the relationship between California pickleweed biomass and soil salinity, I used a linear regression test because soil salinity is a continuous predictor variable and pickleweed biomass is a continuous response variable, and my assumption checks suggested that the assumptions of a linear model were reasonably satisfied.

The test reveals a moderate, positive, and statistically significant relationship between soil salinity and plant biomass, indicating that pickleweed plants at this restoration site tend to grow larger as soil salinity increases (linear regression: β = 2.08, R^2 = 0.286, df = 21, p = 0.0086).

# e. Test implications

The results suggest that pickleweed biomass tends to increase as soil salinity increases, indicating that plants at this site grow larger in areas with higher salinity. While salinity does not explain all of the variation in biomass, the positive and statistically significant relationship suggests that moderately higher salinity conditions may support better pickleweed growth. For planting efforts, this means it may be beneficial to prioritize areas of the slough with higher soil salinity when selecting locations for new pickleweed plantings.

# f. Double check your own work
```{r}
cor.test(salinity$salinity_mS_cm, salinity$pickleweed) #running a pearson's correlation test using salinity and pickleweed biomass data.
```
Both the simple linear regression and the Pearson’s correlation test would lead to rejecting the null hypothesis because both produce a statistically significant p-value (p = 0.0086). However, the regression model estimates how the response variable (pickleweed biomass) changes with the predictor variable (soil salinity), giving a slope of about 2.08 g increase in biomass per 1 mS/cm increase in salinity, while the correlation test summarizes the strength and direction of the association between the two variables (r ≈ 0.53). Although they describe the relationship in different ways, both tests indicate that higher soil salinity is associated with greater pickleweed biomass at this site.






